Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-09-02 Manuscript/manuscript.Rmd
intro edits

A key factor in locating TAD boundaries is Hi-C data "resolution". Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. To increase the resolution of Hi-C data requires a quadratic increase in the total sequencing depth [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to define sub-domains and the boundaries that demarcate them. Furthermore, even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-Sequencing experiments used to annotate the genome, which ranges in the tens of base pairs. This resolution gap between 1D and 3D genomic maps can make it difficult to disentangle the role of additional factors related to TAD boundaries, thereby limiting our understanding of epigenomic regulation and its potential link to chromatin architecture. We hypothesize that, by leveraging genomic annotations--defined by high resolution ChIP-Seq data--known to be enriched at TAD boundaries, we may be able more precisely predict the locations of TAD boundaries on the linear genome.

Several methods have been developed to predict TAD boundary location using genomic annotations. However, most ignore key characteristics of 3D genomic data that are detrimental to model performance. A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, the high level of resolution is likely to introduce heavily imbalanced classes created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. *HiCFeat* did not the address class imbalance. Instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations, resulting in a zero inflated model. <!--This, likely being further exacerbated by the class imbalance.--> Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in basepairs between ChIP-seq annotations and genomic bins in TAD boundary prediction.

Therefore, given the many pitfalls encountered by previous methods we set out to establish a unified approach toward TAD boundary prediction. In this study we first developed a computational framework aimed at determining the set of data level characteristics necessary for optimal TAD boundary prediction. We assessed the Matthews Correlation Coefficient (MCC) of 10-fold cross-validated random forest classifiers with different combinations of model inputs consisting of vary data resolutions, re-sampling techniques, and feature engineering steps. We showed that 10 kb resolution data, combined with both random over- and under-sampling, and a feature space composed of distances between ChIP-seq annotations and genomic bins yielded the best model performances for TAD boundary prediction. We showed that our optimally built model was able to identify known architectural proteins CTCF, RAD21, SMC3, and ZNF143 as exhibiting high predictive importance, confirming their biological functionality near TAD boundaries. We then aimed to use our optimally built model to more precisely predict the location of TAD boundaries at base pair resolution. We showed that, compared to TAD boundaries identified by a TAD caller, these top ranked transcription factors were much more enriched near the predicted TAD boundaries identified by our model indicating much more precise locations throughout the linear genome. This improved precision in TAD boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome.

>>>>

A key factor in locating TAD boundaries is Hi-C data "resolution". Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to define sub-domains and the boundaries that demarcate them. Furthermore, even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-Sequencing experiments used to annotate the genome, which ranges in the tens of base pairs. This resolution gap between 1D and 3D genomic maps can make it difficult to disentangle the role of genomic factors related to TAD boundaries, thereby limiting our understanding of epigenomic regulation and its potential link to chromatin architecture. We hypothesize that, by leveraging genomic annotations--defined by high resolution ChIP-Seq data--known to be enriched at TAD boundaries, we may be able more precisely predict the locations of TAD boundaries at higher resolutions.

Several methods have been developed to predict TAD boundary location using genomic annotations. However, most ignore key characteristics of 3D genomic data that are detrimental to model performance. A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, such high level of resolution is likely to introduce heavily imbalanced classes created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. *HiCFeat* did not the address class imbalance. Instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations, resulting in a zero inflated model. <!--This, likely being further exacerbated by the class imbalance.--> Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other re-sampling solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in base pairs between ChIP-seq annotations and genomic bins in TAD boundary prediction, which offer a more spatial measure of association.

Given the many pitfalls encountered by previous methods we set out to establish a unified approach toward TAD boundary prediction. In this study we first developed a computational framework aimed at determining the set of data level characteristics necessary for optimal TAD boundary prediction. We assessed the Matthews Correlation Coefficient (MCC) of 10-fold cross-validated random forest classifiers with different combinations of model inputs consisting of vary data resolutions, re-sampling techniques, and feature engineering steps. We showed that 10 kb resolution data, combined with both random over- and under-sampling, and a feature space composed of distances between ChIP-seq annotations and genomic bins yielded the best model performances for TAD boundary prediction. We showed that our optimally built model was able to identify known architectural proteins CTCF, RAD21, SMC3, and ZNF143 as exhibiting high predictive importance, confirming their biological functionality near TAD boundaries. We then aimed to use our optimally built model to more precisely predict the location of TAD boundaries at base pair resolution. We showed that, compared to TAD boundaries identified by a TAD caller [@durand2016juicebox], these top ranked transcription factors were much more enriched near the predicted TAD boundaries identified by our model indicating much more precise locations throughout the linear genome. This improved precision in TAD boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome.
